Accelerated Inference for Partially Observed Markov Processes using Automatic Differentiation

### 2024-07-17 paper/ms.tex
updates from overleaf

The convergence of IF2 (and so the first stage of IFAD) to a neighborhood of the MLE happens rapidly in practice. 
In the case of the Dhaka cholera model of \cite{king08}, when an aggressive geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used,  \ed{[initial convergence happens  BE MORE PRECISE] initial convergence to within xxx=40?? log-likelihood units of the maximum} occurs within 40 iterations. 
However, even when applying much larger amounts of Monte Carlo effort, IF2 fails to get within five log-likelihood units of the maximum on this problem (Figure~\ref{fig:boxplot}).
These numerical results are consistent with theoretical results for iterated filtering, showing slow convergence close to the maximum \cite{doucet15-if}.
IF2, without assistance from AD, requires additional computationally intensive techniques, such as repeated pruning and profiling, to reach the inferentially critical region within one log unit of the maximum.

@@

\arxiv{}{\vspace*{-2mm}}
$$
\ell(\theta^*) - \ell(\theta_{m+1}) \leq \Big(1-\eta\beta\, \frac{8\gamma}{9c}\Big)\big(\ell(\theta^*)-\ell(\theta_m)\big).
$$
\label{thm:mop-convergence}
\end{thm}
\arxiv{}{\vspace*{-5mm}}
The proof of Theorem~\ref{thm:linear} uses results from randomized numerical linear algebra, building on Theorem~6 of\cite{mahoney16}. 
Details are deferred to \arxiv{Appendix~\ref{appendix:convergence}}{the supplementary information}. 
\ed{TECHNICALLY, IF WE CLAIM THAT A PROOF FOR ALPHA=1 CAN BE GENERALIZED TO ANY ALPHA, WE ARE CLAIMING A PROOF FOR ALL ALPHA, AND SO WE DO NOT NEED TO SAY WE ARE ONLY PROVING IT FOR ALPHA=1. IF WE REALLY ONLY CLAIM TO HAVE SHOWN IT FOR ALPHA=1, THAT SHOULD BE IN THE STATEMENT. IF WE CLAIM THE WHOLE RESULT, PERHAPS WE CAN POSTPONE THE DISCUSSION OF THIS 2-STAGE METHOD TO THE SUPPLEMENT? AT THIS POINT, ``THE REQUIRED EPSILON-APPROXIMATION'' IS NOT DEFINED FOR THE READER, SO IT IS HARD TO COMPREHEND THE ARGUMENT.}
For simplicity, we only prove this for MOP-$1$ and note that the general case follows as the required $\epsilon$-approximation can still be obtained for sufficiently large $\alpha$. 
When $\alpha=1$, the required $J$ is given in \arxiv{Appendix~\ref{appendix:convergence}}{the supplementary information}. 
This result shows that particle estimates for the gradient can enjoy the same linear convergence rate on well-conditioned problems as gradient descent with access to the score.

>>>>

The convergence of IF2 (and so the first stage of IFAD) to a neighborhood of the MLE happens rapidly in practice. 
In the case of the Dhaka cholera model of \cite{king08}, when an aggressive geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used, we require no more than 40 iterations of IF2 to obtain a suitable warm-start of acceptable quality. 
However, even when applying much larger amounts of Monte Carlo effort, IF2 fails to get within seven log-likelihood units of the maximum on this problem (Figure~\ref{fig:boxplot}).
These numerical results are consistent with theoretical results for iterated filtering, showing slow convergence close to the maximum \cite{doucet15-if}.
IF2, without assistance from AD, requires additional computationally intensive techniques, such as repeated pruning and profiling, to reach the inferentially critical region within one log unit of the maximum.

@@

\arxiv{}{\vspace*{-2mm}}
$$
\ell(\theta^*) - \ell(\theta_{m+1}) \leq \Big(1-\eta\beta\, \frac{8\gamma}{9c}\Big)\big(\ell(\theta^*)-\ell(\theta_m)\big).
$$
\label{thm:mop-convergence}
\end{thm}
\arxiv{}{\vspace*{-5mm}}
The proof of Theorem~\ref{thm:linear} uses results from randomized numerical linear algebra, building on Theorem~6 of\cite{mahoney16}. 
Details are deferred to \arxiv{Appendix~\ref{appendix:convergence}}{the supplementary information}. 
This result shows that particle estimates for the gradient can enjoy the same linear convergence rate on well-conditioned problems as gradient descent with access to the score.
